# 🔹 1. Load required R packages

```{r}
library(tidyverse)  # For data manipulation and ggplot2
library(dplyr)      # For wrangling data
library(ggplot2)    # For plotting
library(lipidr)     # Specialized lipidomics tools
library(readr)      # For reading CSV files
```

# 🔹 2. Import your data

```{r}
dir_path = getwd()
dir_path

matrix <- read_csv(paste0(dir_path, "/lipidomics_matrix.csv"))

coldata <- read_csv(paste0(dir_path, "/lipidomics_metadata.csv"))
```

# 🔹 3. Create a Lipidomics Experiment Object

```{r}
# Convert your raw matrix into a LipidomicsExperiment object (lipidr's main structure)
d <- as_lipidomics_experiment(matrix)

# Add sample group info to the experiment
d <- add_sample_annotation(d, coldata)
```

# 🔹 4. Visualize raw data: Total intensity

```{r}
plot_samples(d, type = "boxplot", log = TRUE)
```
# 🔹 5. Supervised Multivariate Analysis (OPLS-DA)

```{r}
# Compare two groups: CAR vs Mock
mvaresults <- mva(
  d,
  method = "OPLS-DA",
  group_col = "Sample_Type",
  groups = c("Mock", "CAR")
)

# Visualize sample separation
plot_mva(mvaresults, color_by = "Sample_Type")

# Show top lipid features that drive group separation
plot_mva_loadings(mvaresults, color_by = "Class", top.n = 3)

# List top 10 driver lipids
top_lipids(mvaresults, top.n = 10)
```

# 🔹 6. Normalize and Transform Your Data

```{r}
# Normalize lipid intensities (PQN) and log2-transform in one step
d_normalized <- normalize_pqn(d, measure = "Area", log = TRUE)

# Mark that it’s normalized and logged
d_normalized <- set_logged(d_normalized, "Area", TRUE)
d_normalized <- set_normalized(d_normalized, "Area", TRUE)
```

# 🔹 7. Differential Expression (CAR vs Mock)

```{r}
# Identify differentially expressed lipids
de_results <- de_analysis(
  data = d_normalized,
  contrast = CAR - Mock,
  measure = "Area"
)

# Check first few results
head(de_results)

# Visualize distribution again post-normalization
plot_samples(d_normalized, type = "boxplot", log = TRUE)
```

# 🔹 8. Basic Volcano Plot (from lipidr)

```{r}
# List statistically significant lipids
significant_molecules(de_results, p.cutoff = 0.05, logFC.cutoff = 1)

# Plot: Log2 Fold Change vs -log10(P-Value)
plot_results_volcano(de_results, show.labels = FALSE)
```

# 🔹 9. Custom Volcano Plot with ggplot2

```{r}
# Define color/size/opacity styles for plotting
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2.5, "down" = 2.5, "ns" = 1.2)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

# Create fold change categories based on P-value and FC thresholds
volcano_data <- de_results %>%
  drop_na(P.Value, logFC) %>%
  mutate(
    FoldChange = 2^(logFC),
    GeneType = case_when(
      FoldChange >= 2 & P.Value <= 0.1 ~ "up",
      FoldChange <= 0.5 & P.Value <= 0.1 ~ "down",
      TRUE ~ "ns"
    )
  )

# Plot the volcano
p <- ggplot(volcano_data, aes(
  x = log2(FoldChange),
  y = -log10(P.Value),
  fill = GeneType,
  size = GeneType,
  alpha = GeneType
)) +
  geom_point(shape = 21, color = "black") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed") +
  geom_vline(xintercept = c(log2(0.5), log2(2)), linetype = "dashed") +
  scale_fill_manual(values = cols, name = "Regulation") +
  scale_size_manual(values = sizes, guide = "none") +
  scale_alpha_manual(values = alphas, guide = "none") +
  labs(
    title = "Differential Lipids: CAR vs Mock",
    x = "Log2 Fold Change",
    y = "-Log10 P-Value"
  ) +
  theme_minimal(base_size = 13)

# Add labels to up/downregulated lipids
p + ggrepel::geom_text_repel(
  data = filter(volcano_data, GeneType %in% c("up", "down")),
  aes(label = Molecule),
  size = 3,
  max.overlaps = 10
)
```



A 'score' for a Low Medium High










# Stacked bar plot of the broader picture GROUPS 



Standard deviation of 30% and see how they look
# Try the plot_trend(two_group)


